\doctitle

### 2023-04-06 phd_thesis.tex
Initiate PhD thesis structure

\documentclass[11pt,letterpaper]{article}

% PRESETS OF PACKAGES AND MATH %
\input{latex-macros/useful_packages}
\input{latex-macros/math_macros}

%%%%% USER CONFIGURATION %%%%%
\newcommand{\userName}{Aleksei Sholokhov}
\newcommand{\userId}{Lesha}
\newcommand{\department}{Department of Applied Mathematics}
\newcommand{\institution}{University of Washington}
\newcommand{\projectNameShort}{Feature Selection for Mixed Models}
\newcommand{\doctitle}{Thesis Proposal}
\newcommand{\projectNameLong}{}

@@

\title{\doctitle}
~\\
{\center
{\large 
{\huge
Covariate Selection Techniques for Linear Mixed-Effects Models with Applications to Population Health Problems }
\\[10mm]
\userName
\\[40mm]
{\center
\textbf{Committee}\\
Aleksandr Aravkin (chair),\\
James Burke, \\
Nathan Kutz, \\
Jonathan Wakefield (GSR)
}
\\[40mm]
{A thesis proposal submitted in partial fulfillment\\ of the requirements for the degree of}
\\[5mm]
{Doctor of Philosophy}
\\[15mm]
\department
\\[5mm]
\institution
}
\\[10mm]
\textit{Last modified: \today}

@@

\newpage
\section*{Notation}

\begin{center}
	\begin{equation}
		\begin{split}
			n & \text{ -- Total number of objects (observations) in a dataset} \\
			m & \text{ -- Total number of groups (clusters, studies) in a dataset} \\
			n_i & \text{ -- Number of objects (observations) in the group } i. \\
			p & \text{ -- Total number of fixed features (covariates, predictors)} \\
			q & \text{ -- Total number of random features (covariates, predictors)} \\
			X_i & \text{ -- Data (design) matrix of fixed features of the group $i$, } X_i \in \R^{n_i \times p} \\
			Z_i & \text{ -- Data (design) matrix of random features of the group $i$, } Z_i \in \R^{n_i \times q} \\
			Y_i & \text{ -- Target variable for the group } i \\
			\beta & \text{ -- Vector of fixed (mean) coefficients in a model, } \beta \in \R^{p} \\
			\Gamma & \text{ -- Covariance matrix of random effects in a model, } \Gamma \in \R^{q\times q} \\
			\gamma & \text{ -- Vector of variances of random effects when random effects are independent: } \Gamma = \diag(\gamma), \ \gamma \in \R^{q} \\
			u_i & \text{ -- Random effects for the group $i$, } u_i \in \R^q, \ u_i \sim \NN(0, \Gamma) \\
			\Lambda_i & \text{ -- Covariance matrix for measurement (observation) errors for the group $i$} \\
			\sigma_i & \text{ -- Vector of variances for measurement (observation) errors for the group $i$ when } \Lambda_i = \diag(\sigma_i) \\
			\varepsilon_i & \text{ -- Vector of measurement (observation) errors for the group $i$, } \varepsilon_i \sim \NN(0, \Lambda_i) \\
			\Omega_i & \text{ -- Covariance matrix of within-cluster observations, } \Omega_i = \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i, \ \Omega_i \in \R^{n_i \times n_i} \\
			k & \text{ -- Number of non-zero mean coefficients allowed in a model} \\
			s & \text{ -- Number of non-zero variances of random effects allowed in a model} \\
			\tbeta & \text{ -- Sparse counterpart of $\beta$, } \tbeta \in \R^p, \ \|\tbeta\|_0 = k \\
			\tgamma & \text{ -- Sparse counterpart of $\gamma$, } \tgamma \in \R^q, \ \|\tgamma\|_0 = s \\
			\LL(u_i\ |\ \beta, \gamma) & \text{ -- Negative log-likelihood of random effects of a linear mixed model}\\
			\LL_{ML}(\beta, \gamma) & \text{ -- Negative marginalized log-likelihood of parameters of a linear mixed model} \\
			\LL_{REML}(\beta, \gamma) & \text{ -- Negative marginalized restricted log-likelihood (REML) of a linear mixed model}\\
			\LL_{r}(\beta, \gamma) & \text{ -- Relaxed marginalized log-likelihood of a linear mixed model} \\
			a \circ b & \text{ -- Element-wise (Hadamard) product of $a$ and $b$} \\
		\end{split}
	\end{equation}
\end{center}

\newpage

\section{Introduction}
Linear Mixed-Effects Models are widely used technique for modeling clustered data, such as cohort studies, longitudinal data analysis, and meta-research. A key step in these types of works is feature selection, which goal is choosing the most useful predictors from a large set of possible covariates. In the next section we introduce the reader to the concept of linear mixed-effect models and provide a literature review on the most commonly used optimization methods for finding estimators for them, such as Newton method, EM-algorithm, and Interior Point Method. Section \ref{ch:feature_selection_for_mixed_models} gives reader an overview of existing covariates selection methods, such as subset selection, shrinkage penalties, and AIC/BIC-based methods. Section \ref{ch:relax_and_split_thesis_work} introduces Relax-and-Split approach to feature selection, outlines the proposed algorithm, and discusses optimization subroutines involved in it. In the Section \ref{ch:experiments_on_synthetic_data} we test our approach in two simulation setups, with the goals to compare the proposed algorithm to its competitors (Section \ref{ch:comparison_to_other_algo}) and to show its applicability conditions and current limitations (Section \ref{ch:scalability_experiments}). Section \ref{ch:experiments_on_real_data} shows the application of our approach to two real-world problems: the estimation of burden of anxiety and depression disorders as results of bullying (Section \ref{ch:bullying}), and the covariates selection for modeling the contact rate for COVID-19 (Section \ref{ch:covid}). Section \ref{ch:future_work} discusses shortcomings of the current version of the algorithm and outlines theoretical questions and practical improvements which we plan address as a part of the thesis work.

\label{ch:literature_review}
\subsection{Linear Mixed-Effects Models}
Mixed-effect regression models are a particular kind of regression models for clustered data. They describe the relationship between an outcome variable and its predictors in a setting where the observations are grouped according to one or many criteria. These models are used in a wide range of applications, such as longitudinal studies for clinical trials, population health problems, risk-outcome studies, and meta-research.

Mathematically, let there be $m$ groups of observations indexed by $i$, with sizes $n_i$ respectively, and the total number of observations to be $n = n_1 + n_2 + \dots + n_m$. Depending on the domain of application, these groups are sometimes referred to as studies or clusters. For each group, we have design matrices $X_i \in \R^{n_i \times p}$, which are called the matrices of fixed features, and $Z_i \in \R^{n_i \times q}$, called matrices of random features, and we observe vectors of outcomes $Y_i \in \R^{n_i}$. Typically, columns of $Z_i$ are a subset of columns $X_i$ but it does not have to be the case in general. Let $X = [X_1, X_2, \dots, \X_m]^T$ and $Z = [Z_1, Z_2, \dots, Z_m]^T$. According to \cite{Patterson1971, Pinheiro2000}, a linear mixed-effects model (LME) is defined as:

\eq{
	\label{eq:lmm_setup}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
	u_i & \sim \NN(0, \Gamma),\quad \Gamma \in \R_{++}^{q\times q} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \R_{++}^{n_i \times n_i}
}

 where $\beta \in \R^p$ is a vector of fixed (mean) covariates and $u_i \in \R^{q}$ are the vectors of unobservable random effects, which are assumed to be distributed normally with zero mean and the unknown covariance matrix $\Gamma$. The observation error's variances $\Lambda_i$ are often considered to be known, or to have a parametric form, such as $\Lambda_i = \sigma I$ where $\sigma$ is unknown. In this work, we consider $\Lambda_i$ to be known, unless stated otherwise.

 If one considers $\omega_i = Z_i u_i + \varepsilon_i$ to be an unknown per-cluster noise vectors, then the model \ref{eq:lmm_setup} can also be viewed as a correlated noise model:

 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i
}

For the sake of conciseness, since $\Omega_i$ is always a function of $\Gamma$, $\Omega_i(\Gamma)$ will further be spelled as just $\Omega_i$, without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.

 When fitting an LME, the goal is to get both the estimation of fixed covariance $\beta$ and the estimation of variance parameters of random effects $\Gamma$. One of the possible approaches is integrating random effects $u_i$ out to get a marginalized posterior distribution which depends only on $\beta$ and $\Gamma$, but not on $u_i$. According to Bayes Theorem:

\eq{
	\label{eq:integrating_re_out_short}
	f_{\beta, \Gamma | X = \hat{X}, Y=\hat{Y}}(\beta, \Gamma) = \frac{f_{\beta, \Gamma}(\beta, \Gamma)}{f_{X, Y}(\hat{X}, \hat{Y})}\int_{v}f_{X, Y | \beta, u = v}(\hat{X}, \hat{Y})f_{u|\Gamma}(v)dv 
}

where ($\hat{X}, \hat{Y}$) is a particular implementation of random variables $(X, Y)$; $f_{\beta, \Gamma | X=\hat{X}, Y=\hat{Y}}(\beta, \Gamma)$ is a posterior distribution of the model's parameters given data $(\hat{X}, \hat{Y})$, also known as the likelihood of the model's parameters; $f_{\beta, \Gamma}(\beta, \Gamma)$ is a joint prior distribution of the model's parameters; $f_{X, Y | \beta, u}(\hat{X}, \hat{Y})$ is likelihood of the data $(\hat{X}, \hat{Y})$ for the model (\ref{eq:lmm_setup}) given $\beta$ and $u_i$ are known; and $f_{u|\Gamma}(u)$ is a conditional distribution of random effects $u_i$ given that $\Gamma$ is known.

Omitting the parameters-independent multiplier $f_{X, Y}(\hat{X}, \hat{Y})$ and taking the negative log-likelihood of (\ref{eq:integrating_re_out_short}), we obtain $\LL_{ML}$:

\eq{
	\label{eq:lmm_objective}
	& \LL_{ML}(\beta, \Gamma)  \equiv -\log(f_{\beta, \Gamma| X=\hat{X}, Y=\hat{Y}}(\beta, \Gamma)) \sim \\ & \sim   \sum_{i = 1}^m \half(y_i - X_i\beta)^T(Z_i\Gamma^{-1}Z_i^T + \Lambda_i)^{-1}(y_i - X_i\beta) + \half\log{\det{\pa{Z_i \Gamma^{-1}Z_i^T + \Lambda_i}}} -\log(f(\beta, \Gamma)) = \\
	& = \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\log{\det{\Omega_i}} -\log(f_{\beta, \Gamma}(\beta, \Gamma))
}

This expression $\LL_{ML}(\beta, \Gamma)$ is known as the negative log likelihood function of a linear mixed-effects model. The term $-\log(f_{\beta, \Gamma}(\beta, \Gamma))$ represents prior knowledge about $\beta$ and $\Gamma$, and is often referred to as a regularization term. As an example, setting $-\log(f_{\beta, \Gamma}(\beta, \Gamma)) = \frac{1}{2\sigma^2_\beta}\|\beta\|_2^2 + \frac{1}{2\sigma^2_\Gamma}\|\Gamma\|_2^2$ would represent our prior belief that $\beta \sim \NN(0, \sigma^2_\beta)$ and $\Gamma \sim \NN(0, \sigma^2_\gamma)$.

The unknown coefficients $\beta$ and $\Gamma$ are to be obtained via solving the following optimization problem:

\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma) \\
	s.t. & \ \Gamma \in \R_{++}^{s\times s}
}

In a simpler case where $\Gamma = \diag(\gamma)$ is diagonal (often referred to as \textit{the diagonal setup}), the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:

\eq{
	\label{eq:ml_lme_diagonal_setup}
	\min_{\beta, \gamma} & \ \LL_{ML}(\beta, \diag(\gamma)) \\
	s.t. & \ \gamma > 0
	}

\begin{tip}
\paragraph{Use of $\Gamma$ versus $\gamma$.} Starting from Chapter \ref{ch:relax_and_split_thesis_work} this work is focused on the case where the matrix $\Gamma$ is diagonal: $\Gamma = \diag(\gamma)$ (the diagonal setup (\ref{eq:ml_lme_diagonal_setup})), so the unknown covariance parameters of the problem are often referred to as just $\gamma$. This simpler case corresponds to the situation when random effects $u_{ij}$ corresponding to different mean effects $\beta_j$ are independent of each other. In this case the positive definiteness constraint is guaranteed to satisfy whenever $\gamma > 0$. Authors made the decision to start with the diagonal case because it covers their immediate needs in applications, and they see the extensions of the algorithm proposed in Chapter \ref{ch:proposed_algorithm} to the non-diagonal case to be a part of future work. This chapter, however, discusses both cases of $\gamma$ and $\Gamma$ to make connections between the general and the particular cases, and to preserve the notation consistency with the referenced works.
\end{tip}

%Let's consider a simpler case when $\Gamma$ is set to be a diagonal matrix $\diag(\gamma)$, which  The case of $\gamma_j = 0$ for some $j$ is related to $u_i$ being drawn from a degenerate distribution with the variances for $j$'th component to be zero. It implies that random, group dependent, part for $j$'th covariate is zero, and so the variability of the respective coefficient over groups does not help the model to explain the variance in the target variable. Since this particular case of some $\gamma_j$ being zero is directly related to selection of random features, we relax the constraint of $\Gamma$ to be non-negative definite instead of positive-definite. It also does not cause any singularities in the loss function since the matrices $\Omega_i$ are invertible and positive-definite for as long as $\Lambda_i$ are positive-definite and $\Gamma$ is non-negative definite.

The expected value of the posterior mean effects $\beta$ given $\Gamma$ has a closed form representation which immediately follows from (\ref{eq:lmm_correlated_noise_setup}):

\eq{
	\label{eq:beta_formula}
	\beta^{(k+1)} = \argmin_{\beta}\LL_{ML}(\beta, \Gamma^{(k)}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i
}

There is, however, no closed-form solution for $\Gamma$ in general case, so it has to be obtained via an optimization subroutine. There are three aspects that make this problem challenging from the optimization point of view:

\begin{enumerate}
	\item \textbf{Non-convexity}: the loss function $\LL_{ML}$ is not convex with respect to $\Gamma$, which can prevent an optimization method from obtaining a global solution.
	\item \textbf{Overfitting}: the matrix $\Gamma$ contains as many as $\frac{q(q-1)}{2}$ free parameters, overfitting of which may result in near-singularities, especially when the sample size is small \cite{Lange1989}. The diagonal setup $\Gamma = \diag(\gamma)$ is still prone to overfitting, although has less freedom ($q$ free parameters) to do so.
	\item \textbf{Positive Constraint}: the requirement of keeping the matrix $\Gamma$ to be positive definite limits the scope of methods which can be used for optimization. The diagonal setup has a box constraint $\gamma > 0$ which also has to be accommodated by the optimization subroutine.
\end{enumerate}

When $\beta^*$ and $\Gamma^*$, the minimizers for (\ref{eq:ml_lme_optimization_setup}), are obtained, the individual random effects $u_i$ can be found through the minimization of conditional likelihood of random effects:

\eq{
\min_{u_i}\LL(u_i|\beta, \gamma) = \min_{u_i}\left(\half u_i^T\Gamma^{-1}u_i + \half (Z_iu_i - Y_i + F_i(\beta))^T\Lambda_i^{-1}(Z_iu_i - Y_i + F_i(\beta))\right),
}

This problem has a closed form solution which was first derived by \cite{Harville1976} as an extension to Gauss-Markov theorem:

 \eq{
	\label{eq:random_effects_formula}
	u_i = (\Gamma^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta).	
}

As pointed by \cite{Harville1974}, $\Gamma^*$ -- the minimizer for the maximal likelihood formulation \ref{eq:ml_lme_optimization_setup} -- is known to be biased downwards because it does not take into account the loss of $k$ degrees of freedom from estimating $\beta^*$ within the same inference procedure. To overcome this effect they proposed a Restricted Maximal Likelihood (REML) formulation where the loss function is based on $n-k$ linearly independent error contrasts:

\eq{
	\label{eq:reml_lmm_objective}
	\LL_{REML}(\beta, \Gamma) = -\half\log\det\left(\sum_{i=1}^m X_i^T(Z_i\Gamma^{-1}Z_i^T + \Lambda_i)^{-1}X_i\right) + \LL_{ML}(\beta, \Gamma)
}

Although different from ML, the REML formulation does not introduce more complexity to optimization: all the solution methods for ML formulation discussed below also have an adaptation to REML.

Two main optimization approaches to solving mixed effects likelihood formulation are used: EM algorithm \cite{Dempster1977} and Newton-Raphson \cite{Lindstrom1988}.

\subsubsection{EM algorithm for Mixed Models}
One more way of thinking about mixed-effects models is considering \textbf{$\{Y_i\}_{i=1}^m$} to be incomplete data and \textbf{$\{Y_i, u_i\}_{i=1}^m$} to be complete data, where $u_i$ is the unobservable part. In this setting, the problem of fitting a mixed model can also be classified as a statistical inference problem in the presence of missing data, and so the methods for solving this kind of setup might be applicable. One example is EM-algorithm -- a general approach to estimating parameters in problems with missing data developed by \cite{Dempster1977}. As later discussed by \cite{Laird1982}, one can view EM-algorithm as optimizing the approximating function $Q(\beta, \Gamma | \hat{\beta}, \hat{\Gamma})$ -- the likelihood of $(\beta, \Gamma)$ given that the missing data $u_i$ comes from a normal distribution $\NN(\hat{\beta}, Z_i\hat{\Gamma} Z_i^T + \Lambda_i)$. EM-algorithm, in its general form, consists of two steps: Expectation step and Minimization step:

\eq{
	\text{E-step: }\quad & Q^{(k)}(\beta, \Gamma | \hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}) := p\left(\beta, \Gamma | \hat{u_i}^{(k)} =  \expectation{u_i|\hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}}\right) \\
	\text{M-step: }\quad & \hat\beta^{(k+1)}, \hat\Gamma^{(k)}  :=  \argmax_{\beta, \Gamma}\ Q(\beta, \Gamma^{k+1} | \hat{\beta}^{(k)}, \hat{\Gamma}^{(k)}) 
}

On the E-step the function $Q(\beta, \Gamma | \hat{\beta}, \hat{\Gamma})$ is constructed using the expected values of missing data given the previous-iteration parameters $\hat{\beta}, \hat{\Gamma}$, and on M-step $Q$-function is minimized, and its minimizers $\hat{\beta^*}, \hat{\Gamma^*}$ are set to be next-iteration parameters. The process continues until convergence.

As discussed in \cite{Laird1987}, one can express the EM-algorithm's iterations in the linear mixed models setup as:

\eq{
	\beta^{(k+1)} & = \argmin_{\beta}\LL(\beta, \Gamma^{(k)}) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}(\Gamma^{(k)})X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}(\Gamma^{(k)})Y_i \\
	u_i^{(k+1)} & = ((\Gamma^{(k)})^{-1} + Z_i^T\Lambda_i^{-1}Z_i)^{-1}Z_i^T\Lambda_i^{-1}(Y_i - X_i\beta^{(k)}) \\
	\Gamma^{(k+1)} & = \frac{1}{m} \sum_{i=1}^m u_i^{(k)} (u_i^{(k)})^T + \Gamma^{(k)}(I - Z_i\Omega_i^{-1}(\Gamma^{(k)})Z_i^T\Gamma^{(k)})	
	}

EM-algorithm has two main advantages: simplicity of iterations and the absence of constraints in optimization subroutines. The later is especially important: EM-algorithm guarantees that if the initial guess $\Gamma^{(0)}$ was positive definite then all $\Gamma^{(k)}$, including, under certain conditions, the final result $\Gamma^*$, will be positive definite. This removes the necessity to handle the constraint $\Gamma \in \R^{s\times s}_{++}$ separately, e.g. via re-parametrization using Cholesky decomposition, which allows to keep the iterations fast and simple. As an apparent consequence, EM algorithm also correctly handles the box constraint in the diagonal setup~(\ref{eq:ml_lme_diagonal_setup}).

The main drawback of EM-algorithm is its slow convergence speed. In the form above, \cite{Dempster1977} showed that the algorithm monotonically converges, with geometric rate, as established by~\cite{Wu1983}, to a local maximal likelihood estimator. The geometric rate is due to the fact that EM-algorithm can be equivalently viewed as a proximal gradient descent on KL-divergence based likelihood. There have been wide effort on both discovering more statistical properties of the algorithm, as well as on improving its convergence rate. For instance, in his recent work \cite{Balakrishnan2017} provided a quantitative characterization of a region such that the EM-algorithm, initialized within that region, is guaranteed to converge to a "sufficiently good", fixed point: the one within statistical precision of the global optimum. There are also works which are empirically improving the convergence speed of EM in terms of computations and time via implementation-level improvements, like \cite{Liu1994, Liu1998}, to name a few.

\subsubsection{Newton Methods for Mixed Models}
\label{section:newton_methods_for_mixed_models}
Newton method as a second-order optimization method for estimating parameters of mixed models was proposed by \cite{Jennrich1986} and later improved by \cite{Lindstrom1988}. The core idea of this approach is to consider $\hat\beta(\Gamma)$ as a function of $\Gamma$ (see Eq. \ref{eq:beta_formula}), and apply Newton method to the resulting formulation:

\eq{
	\label{eq:lmm_loss_only_of_big_gamma}
	\min_{\Gamma}\ & \LL_{ML}(\beta(\Gamma), \Gamma) \\
	\text{s.t.}\ & \Gamma \in \R^{s\times s}_{+}
}

In a general setup when $\Gamma$ is not diagonal, one can treat the positive-definiteness constraint by reparametrizing the problem via modified Cholesky decomposition, as suggested by \cite{Chen2003}:
\eq{
	\label{eq:dlld_decomposition}
	\Gamma = DLL^TD
}
where $D$ is an $s\times s$ diagonal matrix, and $L$ is an $s\times s$ lower-triangular matrix with ones on the main diagonal. Applying this variables transformation to the equation (\ref{eq:lmm_loss_only_of_big_gamma}) we get an unconstrained optimization problem on diagonal elements of $D$ and on the off-diagonal elements of $L$:

\eq{
	\min_{D, L}\ & \LL_{ML}(\hat{\beta}(DLL^TD), DLL^TD) \\ 
	\text{s.t.}\ &  D_{ij} = 0, \ i \neq j \\ 
	& \ L_{ii} = 1,  \  i = 1 \dots s
}

In a simpler case of $\Gamma$ being diagonal ($\Gamma = \diag{\gamma}$), the setup (\ref{eq:lmm_loss_only_of_big_gamma}) simplifies to

\eq{
	\label{eq:lmm_loss_only_of_gamma}
	\min_{\gamma > 0} \LL(\hat{\beta}(\gamma), \gamma)
}
On each iteration, Newton-like family of methods is looking for $\gamma^{(k+1)}$ -- a minimizer for a quadratic approximation of the problem at the point $\gamma^{(k)}$:
\eq{
	\gamma^{(k+1)} = \argmin_{\Delta\gamma} & \LL_{ML}(\gamma^{(k)}) + \nabla_\gamma\LL_{ML}(\gamma^{(k)})^T \Delta\gamma + \Delta\gamma^T\nabla^2_{\gamma\gamma}\LL_{ML}(\gamma^{(k)})\Delta\gamma\\
	\text{ s.t. } & \gamma^{(k)} + \Delta\gamma \geq 0
}

The positivity constraint narrows the scope of optimization techniques which can be used. There is, however, still wide variety of methods to choose from like Truncated Newton Methods with Barriers \cite{Nash2000}, Projection Methods \cite{Bertsekas1982}, or Interior Point methods \cite{Potra2000}, one can refer to \cite{TrustRegionBook} for a detailed overview of these algorithms. The Interior Point method shall be discussed in more details since it's used in Chapter \ref{ch:proposed_algorithm} of this work.

Interior Point (IP) methods \cite{Kojima1991, Nesterov1994, Wright1997} target the Karush-Kuhn-Tucker optimality conditions for the problem (\ref{eq:lmm_loss_only_of_gamma}) directly by introducing ``margins'' -- non-negative axillary variables which transform inequality constraints into equality constraints:

\eq{
	\label{eq:lmm_loss_only_of_gamma}
	\min_{\gamma, d} & \LL(\gamma) \\
	\text{ s.t. } & \gamma - d = 0 \\
	& d \geq 0
}

The positivity constraint for $d$ is then handled via a logarithmic barrier. The Lagrangian and the KKT system for this problem are:

\eq{
	\label{eq:kkt_conditions}
	F_\mu(\gamma, d, v) & = \LL(\gamma) + v^T(\gamma - d) - \mu\sum_{i=1}^s\log(d_i) \\
	\nabla F & = \begin{bmatrix}
		\nabla_{\gamma} F_\mu \\
		\nabla_{v} F_\mu \\
		\nabla_{d} F_\mu 
	\end{bmatrix} = \begin{bmatrix}
		\nabla \LL(\gamma) + v \\
		\gamma - d \\
		v \circ d - \mu\circ 1
	\end{bmatrix} = 0
}

where $v$ is a vector of dual variables, and $\mu$ is a positive hyper parameter which define the steepness of the logarithmic barrier, and $v\circ d$ denotes the element-wise product.
Application of Newton root-finding method to this KKT system gives us the Interior Point method. In particular, the search direction $[\Delta\gamma, \Delta v, \Delta d]$ as the solution of the following system:

\eq{
	\label{eq:ip_system_as_equation}
	\nabla^2F_\mu\begin{bmatrix}
		\Delta \gamma \\
		\Delta v \\
		\Delta d
	\end{bmatrix} = -\nabla F_\mu\begin{bmatrix}
		\Delta \gamma \\
		\Delta v \\
		\Delta d
	\end{bmatrix}
}

The iteration then looks like
\eq{
	\gamma^{(k+1)} & = \gamma^{(k)} + \alpha \Delta\gamma \\
	v^{(k+1)} & = v^{(k)} + \alpha \Delta v \\
	d^{(k+1)} & = d^{(k)} + \alpha \Delta d
}

$\alpha$ is chosen via line search so all $\gamma^{(k+1)}$, $v^{(k+1)}$, and $d^{(k+1)}$ are positive, and a sufficient descent condition is satisfied:
\eq{
	\|\nabla F_\mu(\gamma^{(k+1)}, v^{(k+1)}, d^{(k+1)})\| \leq 0.99\|\nabla F_\mu(\gamma^{(k)}, v^{(k)}, d^{(k)})\| 
}

Since this is a root finding procedure for the equation (\ref{eq:kkt_conditions}), the natural stopping criterion is 
\eq{
\|\nabla F_\mu(\gamma^{(k)}, v^{(k)}, d^{(k)})\| \leq \varepsilon
}
for some sufficiently small $\varepsilon$.

Interior Point method is a second-order method, so, under certain assumptions, it has a quadratic rate convergence region around the optimal point. This leads to better performance in terms of the number of iterations comparing to EM-algorithm. The performance in terms of time, however, strongly depends on the problem's size. It's due to the fact that working with Hessian information requires solving a linear system, which is not needed for first order methods like EM-algorithm.

\subsection{Feature Selection for Mixed Models}
\label{ch:feature_selection_for_mixed_models}
In statistical learning setting, feature selection is defined as a process of selecting, or ranking, the most important predictors in a dataset. The goal is to get an interpretable model without sacrificing prediction quality. Selection process often happens simultaneously with fitting the model, where a predictor is considered to be selected if its respective coefficient in the model is non-zero. If the desired number of coefficients $k$ is given, then  the feature selection problem can be formulated as minimization of a loss function $f(\theta)$, which is  often proportionate to the negative log-likelihood $\log\, p(\theta| X, Y)$, with respect to its parameters $\theta$ subject to a zero-norm constraint:

\eq{
	\label{eq:general_feature_selection_zero_norm}
	&\min_{\theta}  f(\theta) \\
	& \text{ s.t. } \|\theta\|_0 \leq k 
	}

 In this form the problem is equivalent to the Knapsack problem, which is NP-complete. Many techniques based on exhaustive search, also known as subset selection process, were developed, and showed to be effective when the total number of predictors is small (see \cite{Muller2013}). The same setup appears to be intractable in a high-dimension setting due to the exponential growth of the number of subsets to check when $n \to \infty$, however, it's possible to get an approximate solution via relaxation techniques discussed below.

 One way to work with zero-norm constraint in the Eq.\ref{eq:general_feature_selection_zero_norm} is to relax it to a first norm constraint. An example of such approach is LASSO, or least absolute square shrinkage operator, studied by \cite{Tibshirani1996} for the least squares model:

 \eq{
 	\label{eq:lasso_regression}
 	\min_\beta ||y - X\theta||_2^2 + \lambda||\theta||_1
 }

  It combines the small bias of the least square estimator and the interpretability of the subset selection method, producing, however, biased estimates for the large coefficients, according to \cite{Zou2006}.  This approach was later extended to exhibit other desirable properties, such as lesser bias for larger coefficients (SCAD from \cite{Fan2001}) or simultaneous selection for highly correlated predictors (Elastic Net from \cite{Zou2005}), and to accommodate other model setups, including classification ($\ell_1$-SVM from \cite{Bradley1998, Shalev-Shwartz2011, Belyy2018}), clustering (\cite{Jajuga1991}), and reinforcement learning (\cite{Kim2019}), to name a few.

  Feature selection for linear mixed models is generally a more challenging problem than feature selection for linear regression problem. First, in linear regression setting the observation are independent, whereas in mixed-effects setup they're typically dependent. Second, one needs to do simultaneous selection of mean effect variables ($\beta$) and the ones related to their random effects' covariance structure ($\Gamma$), and these groups can have different relative importance. Finally, selecting random features implies that $\Gamma$ will have zero columns and rows, which prevents it being positive-definite. Consequently, selecting variance parameters leads to boundary issues which make both theoretical analysis and numerical computations more challenging.

   The shrinkage operator approach was first adapted to the mixed-effect models setting by \cite{Lan2006} to perform fixed effects selection. Selection of random effects is generally more difficult than selection of fixed effects alone because removing random effect from the model requires elimination of the entire respective row and column of $\Gamma$. This extension was provided by \cite{Krishna2008}, who  performed simultaneous fixed and random effects selection by combining three components: the modified Cholesky decomposition $\Gamma = DLL^TD$, as suggested by \cite{Chen2003}, the Adaptive LASSO penalty, as in \cite{Lan2006}, and EM algorithm (\cite{Dempster1977}). Their penalized Q-function for EM-algorithm is defined as:
  \eq{
  	\label{eq:krishna_lasso}
  	Q(\beta, D,L|\beta^{(k)}, D^{(k)}, L^{(k)}) = & \mathbb{E}_{u_i | \beta^{(k)}, D^{(k)}, L^{(k)}}\left[{\sum_{i=1}^m \|Y_i - X_i\beta - Z_iDLu_i\|^2}\right] + \\ + & \lambda\left(\sum_{j=1}^k\frac{|\beta_j|}{|\hat\beta_j|} + \sum_{j = 1}^s\frac{|D_{ii}|}{|\bar{D}_{ii}|}\right)
  }
   They suggest tuning the regularization parameter $\lambda$ using a BIC criterion. A similar formulation was used by \cite{Lin2013} who proposed a joint selection with two-stage ALASSO. On stage 1 they approximate the shrinkage penalty with local quadratic approximation and use Newton method to select random effects, and then, on stage 2 they select fixed effects via penalized log-likelihood where the covariance variables are fixed.

   Another type of regularization which is used for feature selection is SCAD \cite{Fan2001} which acts on the first derivative of the penalty function $p_\lambda(\theta)$:
   \eq{
   		p'_{\lambda}(|\theta|) = \lambda\left\{\ind{\theta \leq \lambda} +  \frac{(a\lambda - \theta)_+}{(a-1)\lambda}\ind{\theta - \lambda} \right\}
   }
   The adaptation of SCAD penalty to selecting both fixed and random features in linear mixed model setup was developed by \cite{Fan2012}.

  In high-dimensional setting marginal likelihood is often replaced with a hierarchical approximation of it known as h-likelihood (\cite{Lee1996}), which, for a given data $(\hat{X}, \hat{Y})$ is defined as

  \eq{
  \label{eq:h_likelihood}
  \LL_{H}(\beta, \Gamma, u) = -\log f_{X, Y | \beta, \Gamma, u}(\hat{X}, \hat{Y}) - \log f_{u|\Gamma}(u, \Gamma)
  }

>>>>

\documentclass[12pt,a4paper]{book}

% PRESETS OF PACKAGES AND MATH %
\input{latex-macros/useful_packages}
\input{math_macros-jim}

\usepackage{natbib}
% Color 
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\plum}[1]{\textcolor{Plum}{#1}}
\newcommand{\teal}[1]{\textcolor{teal}{#1}}

%%%%% USER CONFIGURATION %%%%%
\newcommand{\userName}{Aleksei Sholokhov}
\newcommand{\userId}{Lesha}
\newcommand{\department}{Department of Applied Mathematics}
\newcommand{\institution}{University of Washington}
\newcommand{\projectNameShort}{Feature Selection for Mixed Models}
\newcommand{\doctitle}{Doctoral Thesis}
\newcommand{\projectNameLong}{}

@@

\title{\doctitle}
~\\
{\center
{\large 
{\huge
Covariate Selection Techniques for Linear Mixed-Effects Models with Applications to Population Health Problems }
\\[10mm]
\userName
\\[40mm]
{\center
\textbf{Committee}\\
Aleksandr Aravkin (co-chair),\\
Nathan Kutz (co-chair), \\
James Burke, \\
Jonathan Wakefield (GSR)
}
\\[40mm]
{A thesis proposal submitted in fulfillment\\ of the requirements for the degree of}
\\[5mm]
{Doctor of Philosophy}
\\[15mm]
\department
\\[5mm]
\institution
}
\\[10mm]
\textit{Last modified: \today}

@@

\listoftodos

\chapter{MSR3}
\include{msr3}

\chapter{PINODE}
\include{pinode}

\input{cs}

\bibliographystyle{apalike}
\bibliography{bibliography}

\clearpage

\appendix
\chapter{Appendix}
\section{Acknowledgement}
\input{appendix/acknowledgements}

\section{Derivatives of Marginalized Log-likelihood for Linear Mixed Models}
\label{appendix:derivatives_of_lmm}
\input{appendix/derivatives_of_lmms}

\section{Description of Datasets}
\subsection{GBD Bullying Data}
\input{appendix/bullying_features_description}
